Power simulation section

### 2023-08-18 docs/manuscript.Rmd
iterate on discussion

<!-- short summary of key take aways from this tutorial -->
Human-AI interaction research requires careful planning and consideration of statistical power to ensure reliable and meaningful results. While heuristics and helper programs can be useful for simple designs and models, they often fall short when more complex and customized simulations are required. The present tutorial presents a specific example on how to run power simulations for human-AI interaction experiments.

@@

In our case study, we have performed power simulations based on a single set of parameter values that reflect our assumptions of an expected effect size.
Instead of extracting this expected effect size from meta-analyses or pilot data which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge which can be considered a vague theoretical model about the data-generating process.
When sample sizes are chosen based on the results of our power simulations, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present.
However, if the true effect is indeed smaller, power will be lower the study might not be sufficiently informative.
A common, more conservative strategy for sample size justification is to perform power analysis for a smallest effect size of interest (SESOI). 
An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero.
For strategies on the even more difficult task of specifying a plausible SESOI as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022].
When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*.
By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect, or at least get an impression how strongly power depends on the size of the true effect.
In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g., analyse power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8).
However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions.
In this setting, sensitivity power analysis would require to manually specify additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question.
Power could then be simulated for several of these scenarios (across different number of subjects and items as considered earlier).

@@

<!-- Outlook -->
Beyond the specifics, we want to outline six avenues regarding the outlook on power simulations in human-AI interaction research:

1.  The need for power simulations in human-AI interaction research: As this work has clearly pointed out, human-AI interaction research often involves intricate designs and complex models that cannot be adequately addressed by heuristics or simple helper programs. Power simulations offer a solution by providing Human-AI interaction researchers with a tailored approach to estimating statistical power. These simulations take into account the specific study design, account for the underlying assumptions, and offer more accurate power estimates.

2.  Managing simulations with discrete predictor variables: Power simulations become more manageable when all predictor variables are discrete (like in the presented case study) and fixed by the study design. This allows Human-AI interaction researchers to focus on simulating outcome variables while avoiding the need to simulate predictor values, which would introduce additional assumptions. By simplifying the simulation process, researchers can obtain reliable power estimates without compromising accuracy.

3.  Teaching power simulation skills to human-AI interaction researchers: The ability to conduct  power simulations is a valuable skill that should be taught to Human-AI interaction researchers. By incorporating this training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. Equipping Human-AI interaction researchers with the knowledge and tools to perform power simulations empowers them to make informed decisions and enhance the rigor of their studies.

4.  Addressing the mismatch in effort perception: There is often a significant disconnect between the perceived effort required to perform power simulations and the actual effort estimated by researchers and collaborators in Human-AI interaction research. Many researchers request power simulations from statisticians or methodological experts without fully comprehending the complexity and time-consuming nature of these simulations. It is crucial to raise awareness about the effort involved in  power simulations to ensure realistic expectations and effective collaboration between researchers and methodological experts.

5.  Recognizing the value of power simulations: Power simulations are not mere technicalities; they are valuable research contributions that deserve recognition in Human-AI interaction research. They offer insights into the reliability and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. The importance of power simulations can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports.

>>>>

<!-- short summary of key take aways from this tutorial -->
Human-AI interaction research requires careful planning and consideration of statistical power to ensure reliable and meaningful results. While heuristics and user-friendly software can be useful for simple designs and models, they often fall short when more complex and customized simulations are required. The present tutorial presents a specific example on how to run power simulations for human-AI interaction experiments.

@@

In our case study, we have performed power simulations based on a single set of parameter values that reflect our assumptions of an expected effect size.
Instead of extracting this expected effect size from meta-analyses or pilot data which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge.
Domain knowledge can be considered a vague theoretical model about the data-generating process that is less formal and can only be accessed by a back-and-forth exchange in which domain experts assess the plausibility of simulated data.
When sample sizes are chosen based on the results of our power simulations, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present.
However, if the true effect is indeed smaller, power will be lower and the study might not be sufficiently informative.
A common, more conservative strategy for sample size justification is to perform power analysis for a smallest effect size of interest (SESOI). 
An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero.
For strategies on the even more difficult task of specifying a plausible SESOI as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022].
When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*.
By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect, or at least get an impression on how strongly power depends on the size of the true effect.
In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g., analyse power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8).
However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions.
In this setting, sensitivity power analysis would require to manually specify additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question.
Power could then be simulated for several of these scenarios (across different number of subjects and items as considered earlier).

@@

<!-- Outlook -->
Beyond the specifics of our concrete case study, we want to outline six avenues regarding the outlook on power simulations in human-AI interaction research:

1.  The need for power simulations in human-AI interaction research: As our case study has clearly pointed out, human-AI interaction research often involves intricate designs and complex models that cannot be adequately addressed by heuristics or user-friendly software. Power simulations offer a solution by providing Human-AI interaction researchers with a tailored approach to estimating statistical power. These simulations take into account the specific study design, account for the underlying assumptions, and offer more accurate power estimates.

2.  Managing simulations with discrete predictor variables: Power simulations become more manageable when all predictor variables are discrete (like in the presented case study) and fixed by the study design. This allows Human-AI interaction researchers to focus on simulating outcome variables while avoiding the need for complex simulations of predictor values, which would introduce additional assumptions. By simplifying the simulation process, researchers can obtain reliable power estimates without compromising accuracy.

3.  Teaching power simulation skills to human-AI interaction researchers: The ability to conduct  power simulations is a valuable skill that should be taught to Human-AI interaction researchers. By incorporating this training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. Equipping Human-AI interaction researchers with the knowledge and tools to perform power simulations empowers them to make informed decisions and enhance the rigor of their studies. The need to reason about how to simulate plausible data which is in line with the research hypothesis while not violating domain expertise on how plausible data should look like might also contribute to planning more insightful studies that can answer more precise research questions.

4.  Addressing the mismatch in effort perception: There is often a significant disconnect between the amount of effort required to perform power analysis and  the perceived effort estimated by researchers and collaborators in Human-AI interaction research. Many researchers request power simulations from statisticians or methodological experts without fully comprehending the complexity and time-consuming nature of these tailored simulations. It is crucial to raise awareness about the effort involved in power simulations to ensure realistic expectations and effective collaboration between researchers and methodological experts.

5.  Recognizing the value of power simulations: Power simulations are not mere technicalities; they are valuable research contributions that deserve recognition in Human-AI interaction research. They offer insights into the robustness and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. The importance of power simulations can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports.
